A Guide to Pre-Processing High-Throughput Animal Tracking Data

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex
add validation ref

First, users should exclude positions outside the spatial bounds of a study area by comparing position coordinates with the range of acceptable coordinates (the bounding box), and removing positions outside them (Fig. 2.a; Listing 1). 
A bounding box filter does not require a geospatial representation such as a shapefile, and can help remove unreliable data from a tracking system that is less accurate beyond a certain range (e.g. ATLAS; Beardsworth et al. \textit{in prep.}).
In some special cases, users may wish to remove positions \textit{inside} a bounding box, either because movement behaviour within an area is not the focus of a study, or because positions recorded within an area are known to be erroneous.
An example of the former is studies seeking to study transit behaviour between features which can be approximated by their bounding boxes. 
Instances of the latter are likely to be system specific, but are known from ATLAS systems (Bijleveld et al. \textit{in prep.}). 
Bounding boxes are by definition rectangular, and users seeking to filter for other geometries, such as a circular or irregularly-shaped study area, need a geometric intersection between their data and a spatial representation of the area of interest (e.g. shapefile, geopackage, or \texttt{sf}-object in \texttt{R}).
The \texttt{atlastools} function \texttt{atl\_filter\_bounds} implements bounding box and explicit spatial filters and accepts X and Y coordinate ranges, an \texttt{sf}-polygon or multi-polygon object \citep{pebesma2018}, or any combination of the three to filter the data (Listing 1).
% When both coordinate ranges and a polygon are provided, the data is first filtered by the ranges and then the polygon.
% The boolean function argument \texttt{remove\_inside} determines whether positions inside the bounds are retained (\texttt{FALSE}) or removed (\texttt{TRUE}).

@@

Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system (Beardsworth et al. \textit{in prep.}).
GPS systems provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
Tracking data can also include indirect indicators of data quality.
For instance, GPS systems' location error may be indicated indirectly by the number of satellites involved in the localisation.
In reverse-GPS systems too the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
% A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
% \begin{linenomath*}
%     \begin{equation*}
%         SD = \sqrt{{VARX} + {VARY} + 2 \times {COVXY}}
%      \end{equation*}
% \end{linenomath*}
Unreliable positions can be removed by filtering on direct or indirect measures of quality using \texttt{atl\_filter\_covariates} (Listing 2).

@@

We applied the pre-processing pipeline using \texttt{atlastools} functions described above to a calibration dataset to verify that the residence patch method could correctly identify known stopping points (see Fig. 6).
We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 (Beardsworth et al. \textit{in prep.}; Bijleveld et al. \textit{in prep.}).
Stops in the calibration track were recorded as waypoints using a handheld GPS device (Garmin Dakota 10) at each stop.
We estimated the real duration of each stop as the time difference between the first and last position recorded within 50m of each waypoint, within a 10 minute window before and after the waypoint timestamp (to avoid biased durations from revisits).
Stops had a median duration of 10.28 minutes (range: 1.75 minutes -- 20 minutes; see Supplementary Material).
We cleaned the data before constructing residence patches by (1) removing a single outlier ($>$ 15 km away), removing unrealistic movement ($\geq$ 15 m/s), smoothing the data ($K$ = 5), and (4) thinning the data by resampling over a 30 second interval.
The cleaning steps retained 37,324 positions (74.45\%), while thinning reduced these to 1,803 positions (4.8\% positions of the smoothed track).
Details and code are provided in the Supplementary Material (see \textsc{Validating the Residence Patch Method with Calibration Data}).

>>>>

First, users should exclude positions outside the spatial bounds of a study area by comparing position coordinates with the range of acceptable coordinates (the bounding box), and removing positions outside them (Fig. 2.a; Listing 1). 
A bounding box filter does not require a geospatial representation such as a shapefile, and can help remove unreliable data from a tracking system that is less accurate beyond a certain range \citep[e.g. in ATLAS systems][]{beardsworth2021}.
In some special cases, users may wish to remove positions \textit{inside} a bounding box, either because movement behaviour within an area is not the focus of a study, or because positions recorded within an area are known to be erroneous.
An example of the former is studies seeking to study transit behaviour between features which can be approximated by their bounding boxes. 
Instances of the latter are likely to be system specific, but are known from ATLAS systems (Bijleveld et al. \textit{in prep.}). 
Bounding boxes are by definition rectangular, and users seeking to filter for other geometries, such as a circular or irregularly-shaped study area, need a geometric intersection between their data and a spatial representation of the area of interest (e.g. shapefile, geopackage, or \texttt{sf}-object in \texttt{R}).
The \texttt{atlastools} function \texttt{atl\_filter\_bounds} implements bounding box and explicit spatial filters and accepts X and Y coordinate ranges, an \texttt{sf}-polygon or multi-polygon object \citep{pebesma2018}, or any combination of the three to filter the data (Listing 1).
% When both coordinate ranges and a polygon are provided, the data is first filtered by the ranges and then the polygon.
% The boolean function argument \texttt{remove\_inside} determines whether positions inside the bounds are retained (\texttt{FALSE}) or removed (\texttt{TRUE}).

@@

Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system \citep{beardsworth2021}.
GPS systems provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
Tracking data can also include indirect indicators of data quality.
For instance, GPS systems' location error may be indicated indirectly by the number of satellites involved in the localisation.
In reverse-GPS systems too the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
% A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
% \begin{linenomath*}
%     \begin{equation*}
%         SD = \sqrt{{VARX} + {VARY} + 2 \times {COVXY}}
%      \end{equation*}
% \end{linenomath*}
Unreliable positions can be removed by filtering on direct or indirect measures of quality using \texttt{atl\_filter\_covariates} (Listing 2).

@@

We applied the pre-processing pipeline using \texttt{atlastools} functions described above to a calibration dataset to verify that the residence patch method could correctly identify known stopping points (see Fig. 6).
We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 \citep[][; Bijleveld et al. \textit{in prep.}]{beardsworth2021}.
Stops in the calibration track were recorded as waypoints using a handheld GPS device (Garmin Dakota 10) at each stop.
We estimated the real duration of each stop as the time difference between the first and last position recorded within 50m of each waypoint, within a 10 minute window before and after the waypoint timestamp (to avoid biased durations from revisits).
Stops had a median duration of 10.28 minutes (range: 1.75 minutes -- 20 minutes; see Supplementary Material).
We cleaned the data before constructing residence patches by (1) removing a single outlier ($>$ 15 km away), removing unrealistic movement ($\geq$ 15 m/s), smoothing the data ($K$ = 5), and (4) thinning the data by resampling over a 30 second interval.
The cleaning steps retained 37,324 positions (74.45\%), while thinning reduced these to 1,803 positions (4.8\% positions of the smoothed track).
Details and code are provided in the Supplementary Material (see \textsc{Validating the Residence Patch Method with Calibration Data}).
